---
title: "tidy-and-wranlge"
author: "andie"
format: html
---

```{r}
#| message: false

library(dplyr)
library(tidyr)
library(readr)

```

# Data Sources

## Read in data

```{r}
#| message: false

catch_original <- read_csv(
  "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

```

## Explore data

```{r}
colnames(catch_original)

head(catch_original)

summary(catch_original)

unique(catch_original$Region)

```

Keyboard shortcut heads up - Windows: Ctrl + Shift + M - for pipe operator

# Cleaning and reshaping

## Using the select function

```{r}
# use select() to keep the columns well actually use

catch_data <- catch_original %>% select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)

# to drop instead of select
catch_data <- catch_original %>% select(-All, -notesRegCode)
```

## check column types

```{r}
# shows us the data type in the columns
glimpse(catch_data)
```

## changing column contents with mutate()

```{r}
#change Chinook column to "numeric"
catch_clean <- catch_data %>%
  mutate(Chinook = as.numeric(Chinook))

```

## get location of oddball value and change to NA

```{r}
# why are we getting a warning about coerced NAs?
i <- which(is.na(catch_clean$Chinook))
i

#look at the row in the unaltered data
catch_data[i, ]

# fix with mutate
catch_clean <- catch_data %>%
  mutate(Chinook = if_else(condition = Chinook =="I", 
                    true = "1", 
                    false = Chinook), Chinook = as.numeric(Chinook))
#check the type of the chinook column now make sure it has changed and is dbl now
head(catch_clean)

#check problem row of new cleaned data
catch_clean[i,]

```

## Changing file shape using pivot_longer and pivot_wider

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), 
               names_to = "species", values_to = "catch")
head(catch_long)

catch_wide <- catch_long %>%
  pivot_wider(names_from = species, 
              values_from = catch)
head(catch_wide)
```

```{r}
catch_long <- catch_long %>%
    rename(catch_thousands = catch)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000) %>%
    select(-catch_thousands)

head(catch_long)
```

## summary statistics using group_by and summarize()

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch))

head(mean_region)

```

### Count by region

```{r}
n_region <- catch_long %>%
    group_by(Region) %>%
    summarize(n = n())

head(n_region)
```

# Filtering rows using filter()

```{r}
sse_catch <- catch_long %>%
    filter(Region == "SSE")
head(sse_catch)

#try filtering where catch column is over 1 million fish
catch_million <- catch_long %>%
  filter(catch > 1000000)
head(catch_million)

chinook_sse <- catch_long %>%
  filter(Region == "SSE" & species == "Chinook")
head(chinook_sse)
```

## Sorting data with arrange()

```{r}
mean_region <- catch_long %>%
  group_by(Region) %>%
  summarise(mean_catch = mean(catch)) %>%
  arrange(mean_catch)
head(mean_region)

# use desc()
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(desc(mean_catch))

head(mean_region)
```

## Combining datasets together with join()

```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)

catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))
head(catch_joined)
```

## alterntive method for andling key column
```{r}
region_defs <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>%
  select(code, mgmtArea) %>%
  rename(Region = code, Region_Name = mgmtArea)

catch_joined <- left_join(catch_long, region_defs, by = c("Region"))
head(catch_joined)
```

## split and combine values in columns
```{r}
sites_df <- data.frame(site = c("HAW-101", "Haw-103", "OAH-320"))
head(sites_df) 

split_df<- sites_df %>%
  separate(site, into = c("island", "site_number"), sep = "-")
head(split_df)

```

```{r}
dates_df <- data.frame(year = c("1930", "1930", "1932"),
                       month = c("12", "12", "12"), 
                       day = c("14", "15", "16"))
head(dates_df)

dates_unite <- dates_df %>%
  unite(col = date, year, month, day, sep = "-")
head(dates_unite)

```

```{r}
mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>%
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>%
  mutate(Chinook = as.numeric(Chinook)) %>%
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch * 1000) %>%
  group_by(Region) %>%
  summarise(mean_catch = mean(catch)) %>%
  arrange(desc(mean_catch))

```

## save
```{r}

```

