---
title: "Introduction to Quarto"
format: html
bibliography: references.bib
---

## Quarto syntax practice:

#### formatting:

\-*itals* italicized

\- **BOLD** bolded

\- [linked phrase](https://calendar.google.com/calendar/u/0/r/week?tab=rc) takes to google cal!

\- image

\- analysis analysis analysis oops I need a reference go to visual \> tab subheader \> insert citation... result: [@oke2020]

\- need a table visual \> table tab:

| Col1 | Col2 | Col3 |
|------|------|------|
|      |      |      |
|      |      |      |
|      |      |      |

# 2.7 Practice: Literate Analysis with Ocean Water Samples

## 1. Introduction

#### a. About the data

-   Water sample data from Arctic Ocean
-   Data download from the Arctic Data Center: [Tweedie ADC data](http://doi.org/10.18739/A25T3FZ8X)

#### b. Setup

```{r}
#| warning: false

library(readr)
library(here)

```

#### c. Read in data

```{r}
#| warning: false
bg_chem <- read_csv(here("data/BGchem2008data.csv"))

```

## 2. Analysis

#### a. Calculate summary statistics

```{r}
#| eval: false
#| echo: true

#print column names
colnames(bg_chem)

# show data structure
str(bg_chem)

# show first 6 lines for understanding
head(bg_chem)

# summaries by column
summary(bg_chem)

# unique alue in a column
unique(bg_chem$Date)

```

#### b. Calculate means for Redfield ratio

```{r}
# get means of involved nutrients
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
ammn <- mean(bg_chem$NH4)
phosph <- mean(bg_chem$P)



```


#### c. Calculate Redfield ratio

```{r}
#calculate ratio
ratio <- (nitrate + nitrite + ammn) / phosph
```

## 3. Conclusion
:) Here is the mean ratio we calculated: **`r round(ratio, 2)`**
