---
title: "Spatial Data lesson"
format: html
---

# Setup

### install packages

```{r}
library(readr)
library(here)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
library(dplyr)
```

### download and load data

```{r}
knb_url <- "https://dev.nceas.ucsb.edu/knb/d1/mn/v2/object/urn%3Auuid%3Aaceaecb2-1ce0-4d41-a839-d3607d32bb58"

# download zip to data folder
download.file(url = knb_url, destfile = '../data/shapefile_demo_data.zip')

# unzip data
unzip('../data/shapefile_demo_data.zip', exdir = '../data')

# remove .zip file
file.remove('../data/shapefile_demo_data.zip')
```

### explore data

```{r}
ak_region <- read_sf('../data/ak_regions_simp.shp')

plot(ak_region)

class(ak_region)
head(ak_region)
glimpse(ak_region)
```

### check coordinate system of our data

```{r}
st_crs(ak_region)
```

### change CRS with st_transform() to Alaska ALbers projection (3338)

```{r}
ak_region_3338 <- ak_region %>%
  st_transform(crs = 3338)

st_crs(ak_region_3338)

plot(ak_region_3338)
```

# Explore synergy between sf & dplyr packages

### select()

```{r}
# return names of all the columns
colnames(ak_region_3338)

# try selecting a column > it will still include the geometry column !
ak_region_3338 %>%
  select(region)
```

### filter()

```{r}
unique(ak_region_3338$region)

ak_region_3338 %>%
  filter(region == "Southeast")
```

### spatial Joins

\*How many people live in each region?

```{r}
#read in population data
popDF <- read_csv("../data/alaska_population.csv")

# read in as sf DF in WGS (4326) 
pop_4326 <- st_as_sf(popDF, 
                     coords = c("lng", "lat"), 
                     crs = 4326, 
                     remove = F)

class(pop_4326)
head(pop_4326)

```

### joining with st_join()

```{r}
# transform pop data to Alaska Albers
pop_3338 <- st_transform(pop_4326, crs = 3338)

pop_joined_sf <- st_join(pop_3338, 
                         ak_region_3338, 
                         join = st_within)
head(pop_joined_sf)
```

### calculate total population using group_by() and summarise()

```{r}
pop_region <- pop_joined_sf %>%
  as.data.frame() %>% # will remove geom column, sf to df
  group_by(region) %>%
  summarise(total_pop = sum(population))

head(pop_region)
class(pop_region)
```

### join total pops back into the shapefiles

```{r}
tpop_regions <- left_join(ak_region_3338, pop_region, by = "region")

#plot to check
plot(tpop_regions["total_pop"])
```

### Calculate the population by management area

```{r}
pop_mgt_338 <- tpop_regions %>% 
  group_by(mgmt_area) %>%
  summarise(total_pop = sum(total_pop), do_union = F)
plot(pop_mgt_338["total_pop"])
```

### saving spatial objet with write_sf()

```{r}
write_sf(pop_mgt_338, "../data/ak_regions_populations.shp")
```

# Plot with ggplot2

### plot by region total population

```{r}
ggplot(tpop_regions) + 
  geom_sf(aes(fill = total_pop)) +
  labs(fill= "Total Population")  +
  scale_fill_continuous(low="khaki", high = "firebrick", labels = comma)

```

### add rivers of alaska

```{r}
rivers_3338 <- read_sf("../data/ak_rivers_simp.shp")
st_crs(rivers_3338)

ggplot() + 
  geom_sf(data = tpop_regions, aes(fill=total_pop)) +
    geom_sf(data = pop_3338, 
            size = 0.5) +
    geom_sf(data = rivers_3338,
            aes(linewidth = StrOrder)) + # strahler stream order(larger number = larger river)
    scale_linewidth(range = c(0.05, 0.5),
                    guide = "none") +
    labs(title = "Total Population by Alaska Region",
         fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw()
```

# Adding base maps with ggspatial

base maps use psuedo-marcator (ESPG 3857)

```{r}
pop_3857 <- st_transform(pop_3338, crs = 3857)

ggplot(data = pop_3857) + 
  ggspatial::annotation_map_tile(type = "osm", zoom = 4, progress = "none") +
  geom_sf(aes(color = population), fill = NA) + 
  scale_color_continuous(low = "darkkhaki", 
                         high = "firebrick", 
                         labels = comma)
```

### sf objects with leaflet

have to give it both a projection (like Alaska Albers), and that your shapefile must use a geographic coordinate system

```{r}
# first define a leaflet projection for Alaska Albers, and save it as a variable to use later.
espg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS", # which jaa projection library to use
    code = "EPSG:3338", # projection code
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7))

```

### st_transform() to get unprojected coordinate system (WGS 84)

```{r}
pop_region_4326 <- tpop_regions %>%
  st_transform(crs = 4326)
```

### make leaflet map, which are interactible

```{r}
m <- leaflet(options = leafletOptions(crs = espg3338)) %>%
  addPolygons(data = pop_region_4326, 
              fillColor = "gray", 
              weight = 1)
m
```

### make a nicer map

```{r}
pal <- colorNumeric(palette = "Reds", domain = pop_region_4326$total_pop)

m <- leaflet(options = leafletOptions(crs = espg3338)) %>%
addPolygons(
        data = pop_region_4326,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1,
        label = ~ region
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_region_4326$total_pop),
        title = "Total Population")

m
```

### we can even add cities and labels showing their populations

```{r}
pal <- colorNumeric(palette = "Reds", domain = pop_region_4326$total_pop)

m <- leaflet(options = leafletOptions(crs = espg3338)) %>%
addPolygons(
        data = pop_region_4326,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1,
        label = ~ region
    )  %>%
    addCircleMarkers(
        data = pop_4326,
        lat = ~ lat,
        lng = ~ lng,
        radius = ~ log(population / 500),
        # arbitrary scaling
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~ paste0(pop_4326$city, ", population ", comma(pop_4326$population))
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_region_4326$total_pop),
        title = "Total Population")

m
```
